Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-15 paper/sections/conclusions.tex
[ paper ] cosmetics.
[ paper ] removed file field from ref.bib.

Despite the fact, our logic domain is more straightforward than
first-order or high-order logic where there are many
proof-reconstruction tools (see a list of some of them in the
related work, section \ref{sec:related-work}), we found some issues
around the reconstruction that show this task is far from trivial.
Depending  on the systems involved in the translation, some
difficulties can arise. We listed some of them in section
\ref{sec:metis-language-and-proofs} and in the section
\ref{sec:proof-reconstruction}.

>>>>

Despite the fact, our logic domain is more straightforward than
first-order or high-order logic where there are many
proof-reconstruction tools (see a list of some of them in the
related work, Section \ref{sec:related-work}), we found some issues
around the reconstruction that show this task is far from trivial.
Depending  on the systems involved in the translation, some
difficulties can arise. We listed some of them in Section
\ref{sec:metis-language-and-proofs} and in Section
\ref{sec:proof-reconstruction}.

### 2017-09-15 paper/sections/introduction.tex
[ paper ] removed file field from ref.bib.

Proof-reconstruction is a hard labor since it depends on the
integration of two complex systems. On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories. A \SAT solver (e.g., \name{zChaff}
\cite{Moskewicz2001} and \name{MiniSat}~\cite{Een2004}) to prove
unsatisfiability of \CNF formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and \name{DepQBF}
\cite{Lonsing2017}) to prove satisfiability and invalidity of
quantified Boolean formulas, a \SMT solver (e.g \name{CVC4}
\cite{Barrett2011}, \name{veriT}~\cite{bouton2009}, and \name{Z3}
\cite{DeMoura2008}) to prove unsatisfiability of formulas from
first-order logic with theories, and a prover for validity of
formulas from first-order logic with equality (e.g., \name{E}
\cite{Schulz:AICOM-2002}, \name{leanCoP}~\cite{Otten2008}, \Metis
\cite{hurd2003first}, \name{SPASS} ~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II} \cite{Benzmuller2008} and \name{Satallax}
\cite{Brown2012}) or intuitionistic logic (e.g., \name{ileanCoP}
\cite{Otten2008}, \name{JProver} \cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam}, \name{Coq}
\cite{coqteam}, \name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems like
operators, logic variables, axioms, and theorems.

A proof-reconstruction tool provides such an integration translating
the reply delivered by the prover into the formalism of the proof
assistant. Because the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the
target, the reconstruction turns out in a ``reverse engineering''
task. Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the specification
logic in the \ITP. To begin with the proof-reconstruction, it is
necessary to have a proof-object from the \ATP in a consistent
format to work with, that is, a full script describing step-by-step
with exhaustive details and without ambiguities, the derivation to
get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g., \name{CVC4}
\cite{Barrett2011}, \name{SPASS}, and \name{Waldmeister}
\cite{hillenbrand1997}) and a little bit less replied with a proof
in a file format like \TSTP~\cite{sutcliffe2004tstp} (e.g.,
\name{E}, \Metis, \name{Vampire}, and \name{Z3}), \len{LFSC}
\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been
implemented for proof-reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013}, \cite{Bohme2011}, \cite{Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related
work, section \ref{sec:related-work}.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows.
We briefly mention notation and syntax of the classical propositional
logic in Section \ref{sec:preliminaries}.
In section \ref{sec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
We listed some conclusions and future work in Section
\ref{sec:conclusions}. At the end, we list some related works in
proof reconstruction in Section \ref{sec:related-work}.
The appendix \textsc{A} shows a complete example of the proof
reconstruction workflow.

>>>>

Proof-reconstruction is a hard labor since it depends on the
integration of two complex systems. On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories.
A \SAT solver (e.g., \name{zChaff}~\cite{Moskewicz2001} and
\name{MiniSat}~\cite{Een2004}) to prove unsatisfiability of \CNF
formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and
\name{DepQBF}~\cite{Lonsing2017}) to prove satisfiability and
invalidity of quantified Boolean formulas, a \SMT solver
(e.g \name{CVC4}~\cite{Barrett2011},\name{veriT}~\cite{bouton2009},
and \name{Z3}~\cite{DeMoura2008}) to prove unsatisfiability of
formulas from first-order logic with theories, and a prover for
validity of formulas from first-order logic with equality
(e.g.,\name{E}~\cite{Schulz:AICOM-2002},
\name{leanCoP}~\cite{Otten2008},
\Metis~\cite{hurd2003first},
\name{SPASS}~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II}~\cite{Benzmuller2008} and
\name{Satallax}~\cite{Brown2012}) or intuitionistic logic (e.g.,
\name{ileanCoP}~\cite{Otten2008},
\name{JProver}~\cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam},
\name{Coq}~\cite{coqteam},
\name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems like
operators, logic variables, axioms, and theorems.

A proof-reconstruction tool provides such an integration translating
the reply delivered by the prover into the formalism of the proof
assistant. Because the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the
target, the reconstruction turns out in a ``reverse engineering''
task. Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the specification
logic in the \ITP. To begin with the proof-reconstruction, it is
necessary to have a proof-object from the \ATP in a consistent
format to work with, that is, a full script describing step-by-step
with exhaustive details and without ambiguities, the derivation to
get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g.,
\name{CVC4}~\cite{Barrett2011}, \name{SPASS}, and
\name{Waldmeister}~\cite{hillenbrand1997}) and a little bit less replied with a proof
in a file format like \TSTP~\cite{sutcliffe2004tstp} (e.g.,
\name{E}, \Metis, \name{Vampire}, and \name{Z3}),
\len{LFSC}~\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been
implemented for proof-reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013}, \cite{Bohme2011}, \cite{Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related
work, Section \ref{sec:related-work}.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows.
We briefly mention notation and syntax of the classical propositional
logic in Section \ref{sec:preliminaries}.
In Section \ref{sec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
We listed some conclusions and future work in Section
\ref{sec:conclusions}. At the end, we list some related works in
proof reconstruction in Section \ref{sec:related-work}.
The appendix \textsc{A} shows a complete example of the proof
reconstruction workflow.

### 2017-09-15 paper/sections/metis-language-and-proofs.tex
[ paper ] cosmetics.
[ paper ] removed file field from ref.bib.

In this part, our main purpose consists of giving a response to
these rules looking at the \Metis' source code while inspecting a
representative set of solutions of \CPL problems
\cite{Prieto-Cubides2017} generated by \Metis.
This is an attempt to fulfill the \Metis's documentation, by
describing each rule from our perspective. Later on, in section
\ref{sec:proof-reconstruction}, we present a formal description of
such rules.

@@

\textit{Splitting}. To prove a goal, \Metis splits the goal into
disjoint cases. This process produces a list of new subgoals, the
conjunction of these subgoals implies the goal as we show later in
subsection~\ref{ssec:emulating-inferences}. Then, a proof of the
goal becomes in smaller proofs, one refutation for each subgoal.
These subgoals are introduced in the \TSTP derivation with the \strip
inference rule.

>>>>

In this part, our main purpose consists of giving a response to
these rules looking at the \Metis' source code while inspecting a
representative set of solutions of \CPL problems
\cite{Prieto-Cubides2017} generated by \Metis.
This is an attempt to fulfill the \Metis's documentation, by
describing each rule from our perspective. Later on, in Section
\ref{sec:proof-reconstruction}, we present a formal description of
such rules.

@@

\textit{Splitting}. To prove a goal, \Metis splits the goal into
disjoint cases. This process produces a list of new subgoals, the
conjunction of these subgoals implies the goal as we show later in
Subsection~\ref{ssec:emulating-inferences}. Then, a proof of the
goal becomes in smaller proofs, one refutation for each subgoal.
These subgoals are introduced in the \TSTP derivation with the \strip
inference rule.

### 2017-09-15 paper/sections/proof-reconstruction.tex
[ paper ] removed file field from ref.bib.

\newcommand{\rank}{\name{rank}}

@@

\begin{equation*}
\label{eq:rank-definition}
\rank(φ)=
\begin{cases}
0 &\text{for atomic }φ, \\
\rank(φ₁) + c_{¬}  & φ ≡ \neg φ₁, \\
\max(\rank(φ₁),\rank(φ₂)) + c_{\square}
  & φ ≡ φ₁\,\square\, φ₂.
\end{cases}
\end{equation*}

@@

We reformulate \mysplit function to match with the
new definition after apply the process described above in
subsection~\ref{ssec:small-structured-recursion}, since
in the recursion calls of \mysplit in Eq.~\ref{eq:split-definition},
the negations, $(¬ φ_{1,2})$ raise a terminating problem.

@@

\begin{proof} We now proceed by induction on the structure of the
formula φ in the definition of Eq.~\ref{eq:split-definition}.
For a complete proof, we refer the reader to \cite{AgdaMetis}.\\[3mm]

>>>>

@@

\begin{equation*}
\label{eq:rank-definition}
rank(φ)=
\begin{cases}
0 &\text{for atomic }φ, \\
rank(φ₁) + c_{¬}  & φ ≡ \neg φ₁, \\
\max{(rank(φ₁),rank(φ₂)} + c_{\square}
  & φ ≡ φ₁\,\square\, φ₂.
\end{cases}
\end{equation*}

@@

We reformulate \mysplit function to match with the
new definition after apply the process described above in
Subsection~\ref{ssec:small-structured-recursion}, since
in the recursion calls of \mysplit in Eq.~\ref{eq:split-definition},
the negations, $(¬ φ_{1,2})$ raise a terminating problem.

@@

\begin{proof} We now proceed by induction on the structure of the
formula φ in the definition of Eq.~\ref{eq:split-definition}.
For a complete proof, we refer the reader to~\cite{AgdaMetis}.\\[3mm]

### 2017-09-15 paper/sections/related-work.tex
[ paper ] cosmetics.
[ paper ] removed file field from ref.bib.

\name{Sledgehammer} is a tool for \name{Isabelle} proof assistant
\cite{paulson1994isabelle} that provides a full integration of
automatic theorem provers including \ATPs (see, for example,
\cite{meng2006automation}, \cite{blanchette2013extending} and
\cite{Fleury2014}) and \SMT solvers (see, for instance,
\cite{hurlin07practical}, \cite{bohme2010},
\cite{blanchette2013extending}, and \cite{Fleury2014}) with
\name{Isabelle/HOL} \cite{nipkow2002isabelle}, the specialization of
\name{Isabelle} for higher-order logic. Sultana, Benzm{\"{u}}ller,
and Paulson~\cite{Een2004} integrates \name{Leo-II} and
\name{Satallax}, two theorem provers for high-order logic with
\name{Isabelle/HOL} proposing a modular proof-reconstruction work
flow.

\name{SMTCoq}~\cite{armand2011,Ekici2017} is a tool for the
\name{Coq} proof assistant \cite{coqteam} which provides a certified
checker for proof witness coming from the \SMT solver \name{veriT}
\cite{bouton2009} and adds a new tactic named verit, that calls
\name{veriT} on any \name{Coq} goal. In \cite{bezem2002automated},
given a fixed but arbitrary first-order signature, the authors
transform a proof produced by the first-order automatic theorem
prover \name{Bliksem} \cite{deNivelle2003} in a \name{Coq} proof
term.

Hurd~\cite{Hurd1999} integrates the first-order resolution prover
\name{Gandalf} with \name{HOL}~\cite{norrish2007hol}, a high-order
theorem prover, following an \abbre{LCF} model implementing the
tactic \verb!GANDALF_TAC!. For \name{HOL Light}, a version of
\name{HOL} but with a simpler logic core, the \SMT solver \name{CVC4}
was integrated, and Kaliszyk and Urban~\cite{kaliszyk2013}
reconstruct proofs from different ATPs with their \name{PRocH} tool,
replaying the detailed inference steps from the ATPs with internal
inference methods implemented in \name{HOL Light}.

@@

\name{Waldmeister} is an automatic theorem prover for unit
equational logic \cite{hillenbrand1997}. Foster and Struth
\cite{foster2011integrating} integrate \name{Waldmeister} into \Agda
\cite{agdateam}. This integration requires a proof-reconstruction
step, but authors' approach is restricted to pure equational logic
--also called identity theory~\cite{humberstone2011}-- that is,
first-order logic with equality but no other predicate symbols and
no functions symbols~\cite{appel1959}.

% Kanso and Setzer~\cite{kanso2016light} integrate \name{Z3} in \Agda
% but they also integrate the propositional fragment of the \name{E} prover in
% \cite{Kanso2012}. They cataloged these two integration as
% \emph{Oracle and Reflection} and \emph{Oracle and Justification}, respectively.
% Their integration with the \name{E} prover is the most related work with our
% approach found at the moment while we write this document. Besides the use
% of the same \ITP, crucial differences easily arise like decidability of
% propositional logic using semantics that we do not take into account for the
% reason we describe later on in section~\ref{sec:conclusions}.

>>>>

\name{Sledgehammer} is a tool for \name{Isabelle} proof assistant
\cite{paulson1994isabelle} that provides a full integration of
automatic theorem provers including \ATPs (see, for example,
\cite{meng2006automation}, \cite{blanchette2013extending} and
\cite{Fleury2014}) and \SMT solvers (see, for instance,
\cite{hurlin07practical}, \cite{bohme2010},
\cite{blanchette2013extending}, and \cite{Fleury2014}) with
\name{Isabelle/HOL} \cite{nipkow2002isabelle}, the specialization of
\name{Isabelle} for higher-order logic.
\citeauthor{Een2004}~\cite{Een2004} integrates \name{Leo-II} and
\name{Satallax}, two theorem provers for high-order logic with
\name{Isabelle/HOL} proposing a modular proof-reconstruction work
flow.

\name{SMTCoq}~\cite{armand2011,Ekici2017} is a tool for the
\name{Coq} proof assistant \cite{coqteam} which provides a certified
checker for proof witness coming from the \SMT solver
\name{veriT}~\cite{bouton2009} and adds a new tactic named verit,
that calls \name{veriT} on any \name{Coq} goal.
In \cite{bezem2002automated},
given a fixed but arbitrary first-order signature, the authors
transform a proof produced by the first-order automatic theorem
prover \name{Bliksem} \cite{deNivelle2003} in a \name{Coq} proof
term.

Hurd~\cite{Hurd1999} integrates the first-order resolution prover
\name{Gandalf} with \name{HOL}~\cite{norrish2007hol}, a high-order
theorem prover, following an \abbre{LCF} model implementing the
tactic \verb!GANDALF_TAC!. For \name{HOL Light}, a version of
\name{HOL} but with a simpler logic core, the \SMT solver \name{CVC4}
was integrated, and \citeauthor{kaliszyk2013}~\cite{kaliszyk2013}
reconstruct proofs from different ATPs with their \name{PRocH} tool,
replaying the detailed inference steps from the ATPs with internal
inference methods implemented in \name{HOL Light}.

@@

\name{Waldmeister} is an automatic theorem prover for unit
equational logic \cite{hillenbrand1997}.
\citeauthor{foster2011integrating}~\cite{foster2011integrating}
integrate \name{Waldmeister} into \Agda
\cite{agdateam}. This integration requires a proof-reconstruction
step, but authors' approach is restricted to pure equational logic
--also called identity theory~\cite{humberstone2011}-- that is,
first-order logic with equality but no other predicate symbols and
no functions symbols~\cite{appel1959}.

% Kanso and Setzer~\cite{kanso2016light} integrate \name{Z3} in \Agda
% but they also integrate the propositional fragment of the \name{E} prover in
% \cite{Kanso2012}. They cataloged these two integration as
% \emph{Oracle and Reflection} and \emph{Oracle and Justification}, respectively.
% Their integration with the \name{E} prover is the most related work with our
% approach found at the moment while we write this document. Besides the use
% of the same \ITP, crucial differences easily arise like decidability of
% propositional logic using semantics that we do not take into account for the
% reason we describe later on in Section~\ref{sec:conclusions}.
